\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-02 paper/ms.tex
words about astero example

We set the mean function $\mu_\bvec{\theta}$ to zero and the remaining
parameters $\bvec{\alpha}$ and their priors are listed in
Table~\ref{tab:example-3-params}.
As with the earlier examples, we start by estimating the MAP parameters using
\project{L-BFGS-B} and initialize \exampleiiinwalkers~walkers by sampling from
an isotropic Gaussian with a standard deviation of $10^{-5}$ centered on the
MAP parameters.
}

We used \emcee\ to sample the joint posterior probability density for the four
parameters in \eq{rot-kernel}, conditioned on two quarters of \kepler\ data,
and find a constraint on the rotation period of $P = \rotationperiod$.
This is in good agreement with the literature value with smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
\etal, submitted) demonstrates that GP inferences are, at a population level,
more reliable than other methods.
The left panels of \Figure{rotation} show a subset of the data used in this
example and the residuals away from the maximum likelihood predictive mean.
The right panel of \Figure{rotation} shows the marginalized posterior
distribution for the rotation period.

@@

\subsection{Asteroseismic oscillations}

The asteroseismic oscillations of thousands of stars were measured using light
curves from the \kepler\ mission \citep{Gilliland:2010, Huber:2011,
Chaplin:2011, Chaplin:2013, Stello:2013} and asteroseismology is a key science
driver for many of the upcoming large scale photometric surveys
\citep{Campante:2016, Rauer:2014, Gould:2015}.
Most asteroseismic analyses have been limited to relatively high
signal-to-noise oscillations.
The standard methods use statistics of the empirical periodogram and cannot
formally propagate the measurement uncertainties to the constraints on
physical parameters, instead relying on bootstrapped uncertainty estimators
\citep{Huber:2009}.
More sophisticated methods that compute the likelihood function in the time
domain scale poorly to state-of-the-art datasets \citep{Brewer:2009,
Corsaro:2014}.

\celerite\ alleviates these problems by providing a physically motivated
probabilistic model that can be evaluated efficiently even for large datasets.
In practice, one would model the star as a mixture of stochastically-driven
simple harmonic oscillators where the amplitudes and frequencies of the
oscillations are computed using a physical model, and evaluate the probability
of the observed time series using a GP where the PSD is a sum of terms given
by \eq{sho-psd}.
This gives us a method of computing the likelihood function for the parameters
of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta \nu$, or
other more fundamental parameters) \emph{conditioned on the observed time
series} in $\mathcal{O}(N)$ operations.
In other words, \celerite\ provides a computationally efficient framework that
can be combined with physically-motivated models of stars and numerical
inference methods to make rigorous probabilistic measurements of asteroseismic
parameters in the time domain.
We expect that this has the potential to push asteroseismic analysis to lower
signal-to-noise datasets and we hope to revisit this idea in a subsequent
paper.

To demonstrate the method, we use a simple heuristic model where the PSD is
given by a mixture of 8 components with amplitudes and frequencies specified
by $\nu_\mathrm{max}$, $\Delta \nu$, and several nuisance parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters. \dfmtodo{A was used for diagonal of
covariance matrix.}
This model could be extended to include more angular modes and
$\nu_\mathrm{max}$ and $\Delta \nu$ could be replaced by the fundamental
physical parameters of the star.

To demonstrate the applicability of this model, we apply it to infer the
asteroseismic parameters of the giant star KIC~11615890, observed by the
\kepler\ Mission.
The goal of this example is to show that, even for a low signal-to-noise
dataset with a short baseline, it is possible to infer asteroseismic
parameters with formal uncertainties that are consistent with the parameters
inferred with a much larger dataset.
Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we measure
$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
compare our results to the results inferred from the full 4 year baseline of
the \kepler\ mission.
Unlike typical giants, KIC~11615890 is a member of a class of stars where the
dipole ($\ell = 1$) oscillation modes are suppressed by strong magnetic fields
in the core \citep{Stello:2016}.
This makes this target simpler for the purposes of this demonstration because
we can neglect the $\ell = 1$ modes and the model proposed above will be an
effective model for the combined signal from the $\ell = 0$ and 2 modes.

For KIC~11615890, the published asteroseismic parameters measured using several
years of \kepler\ observations are \citep{Pinsonneault:2014}
\begin{eqnarray}
    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
\end{eqnarray}
We randomly select a month-long segment of PDC \citep{Stumpe:2012, Smith:2012}
\kepler\ data, initialize our \celeriteterm\ model using a grid search in the
parameter space, and then use \emcee\ \citep{Foreman-Mackey:2013} to sample
the joint posterior density for the full set of parameters.
\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
and $\Delta\nu$ compared to the results from the literature.
This result is consistent within the published error bars and the posterior
constraints are tighter than the published results.
All asteroseismic analyses are known to have substantial systematic and
method-dependent uncertainties \citep{Verner:2011} so further experiments
would be needed to assess the reliability of this method.

This model requires about a minute of computation time to find the maximum
likelihood parameters and then it takes about an hour to run the MCMC sampling
to convergence on a dual-CPU laptop.
This is more computationally intensive than traditional methods of measuring
asteroseismic oscillations, but is is also orders of magnitude cheaper than the
same analysis using another GP solver.
An in-depth discussion of the benefits of rigorous probabilistic inference of
asteroseismic parameters in the time domain is beyond the scope of this paper,
but we hope to revisit this opportunity in the future.

@@

\subsection{Exoplanet transit fitting}

@@

This paper includes data collected by the \kepler\ mission. Funding for the
\kepler\ mission is provided by the NASA Science Mission directorate.
We are grateful to the entire \kepler\ team, past and present.
These data were obtained from the Mikulski Archive for Space Telescopes
(MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST is provided by the NASA Office of Space Science via grant
NNX13AC07G and by other grants and contracts.

>>>>

We set the mean function $\mu_\bvec{\theta}$ to zero and the remaining
parameters $\bvec{\alpha}$ and their priors are listed in
Table~\ref{tab:example-3-params}.
As with the earlier examples, we start by estimating the MAP parameters using
\project{L-BFGS-B} and initialize \exampleiiinwalkers~walkers by sampling from
an isotropic Gaussian with a standard deviation of $10^{-5}$ centered on the
MAP parameters.
The left panels of \Figure{rotation} show a subset of the data used in this
example and the residuals away from the MAP predictive mean.

We run \exampleiiinburn~steps of burn-in, followed by \exampleiiinsteps~steps
of MCMC using \emcee.
We estimate the integrated autocorrelation time of the chain for $\ln P$ and
estimate that we have \exampleiiineff\ independent samples.
These samples give a posterior constraint on the period of $P =
\rotationperiod$ and the marginalized posterior distribution for $P$ is shown
in the right panel of \Figure{rotation}.
This result is in good agreement with the literature value with smaller
uncertainties.
A detailed comparison of GP rotation period measurements and the traditional
methods is beyond the scope of this paper, but forthcoming work (R.~Angus,
\etal, submitted) demonstrates that GP inferences are, at a population level,
more reliable than other methods.
}

@@

\subsection{Example 4: Asteroseismic oscillations}

The asteroseismic oscillations of thousands of stars were measured using light
curves from the \kepler\ Mission \citep{Gilliland:2010, Huber:2011,
Chaplin:2011, Chaplin:2013, Stello:2013} and asteroseismology is a key science
driver for many of the upcoming large scale photometric surveys
\citep{Campante:2016, Rauer:2014, Gould:2015}.
Most asteroseismic analyses have been limited to high signal-to-noise
oscillations because the standard methods use statistics of the empirical
periodogram and cannot formally propagate the measurement uncertainties to the
constraints on physical parameters, instead relying on bootstrapped
uncertainty estimators \citep{Huber:2009}.
More sophisticated methods that compute the likelihood function in the time
domain scale poorly to large survey datasets \citep{Brewer:2009,
Corsaro:2014}.

\celerite\ alleviates these problems by providing a physically motivated
probabilistic model that can be evaluated efficiently even for large datasets.
In practice, we model the star as a mixture of stochastically-driven simple
harmonic oscillators where the amplitudes and frequencies of the oscillations
are computed using a physical model, and evaluate the probability of the
observed time series using a GP where the PSD is a sum of terms given by
\eq{sho-psd}.
This gives us a method for computing the likelihood function for the
parameters of the physical model (for example, $\nu_\mathrm{max}$ and $\Delta
\nu$, or other more fundamental parameters) \emph{conditioned on the observed
time series} in $\mathcal{O}(N)$ operations.
In other words, \celerite\ provides a computationally efficient framework that
can be combined with physically-motivated models of stars and numerical
inference methods to make rigorous probabilistic measurements of asteroseismic
parameters in the time domain.
This has the potential to push asteroseismic analysis to lower signal-to-noise
datasets and we hope to revisit this idea in a future paper.

To demonstrate the method, we use a simple heuristic model where the PSD is
given by a mixture of 8 components with amplitudes and frequencies specified
by $\nu_\mathrm{max}$, $\Delta \nu$, and some nuisance parameters.
The first term is used to capture the granulation ``background''
\citep{Kallinger:2014} using \eq{granulation-psd} with two free parameters
$S_g$ and $\omega_g$.
The remaining  7 terms are given by \eq{sho-psd} where $Q$ is a nuisance
parameter shared between terms and the frequencies are given by
\begin{eqnarray}
\omega_{0,\,j} = 2\,\pi\,(\nu_\mathrm{max} + j\,\Delta\nu + \epsilon)
\end{eqnarray}
and the amplitudes are given by
\begin{eqnarray}
S_{0,\,j} =
    \frac{A}{Q^2}\,\exp\left(-\frac{[j\,\Delta\nu + \epsilon]^2}{2\,W^2}\right)
\end{eqnarray}
where $j$ is an integer running from $-3$ to 3 and $\epsilon$, $A$, and $W$
are shared nuisance parameters.
\response{Finally, we also fit for the amplitude of the white noise by adding
a parameter $\sigma$ in quadrature with the uncertainties given for the
\kepler\ observations.
All of these parameters and their chosen priors are listed in
Table~\ref{tab:example-4-params}.
As before, these priors are all log-uniform except for $\epsilon$ where we use
a zero-mean normal prior with a broad variance of $1~\mathrm{day}^2$ to break
the degeneracy between $\nu_\mathrm{max}$ and $\epsilon$.
}
To build a more realistic model, this prescription could be extended to
include more angular modes, or $\nu_\mathrm{max}$ and $\Delta \nu$ could be
replaced by the fundamental physical parameters of the star.

To demonstrate the applicability of this model, we apply it to infer the
asteroseismic parameters of the giant star KIC~11615890, observed by the
\kepler\ Mission.
The goal of this example is to show that, even for a low signal-to-noise
dataset with a short baseline, it is possible to infer asteroseismic
parameters with formal uncertainties that are consistent with the parameters
inferred with a much larger dataset.
Looking forward to \tess\ \citep{Ricker:2014,Campante:2016}, we measure
$\nu_\mathrm{max}$ and $\Delta\nu$ using only one month of \kepler\ data and
compare our results to the results inferred from the full 4 year baseline of
the \kepler\ Mission.
For KIC~11615890, the published asteroseismic parameters measured using several
years of \kepler\ observations are \citep{Pinsonneault:2014}
\begin{eqnarray}
    \nu_\mathrm{max} = 171.94 \pm 3.62 \,\mu\mathrm{Hz} \quad\mathrm{and}\quad
    \Delta\nu = 13.28 \pm 0.29 \,\mu\mathrm{Hz} \quad.
\end{eqnarray}

Unlike typical giants, KIC~11615890 is a member of a class of stars where the
dipole ($\ell = 1$) oscillation modes are suppressed by strong magnetic fields
in the core \citep{Stello:2016}.
This makes this target simpler for the purposes of this demonstration because
we can neglect the $\ell = 1$ modes and the model proposed above will be an
effective model for the combined signal from the $\ell = 0$ and 2 modes.

\response{
For this demonstration, we randomly select a month-long segment of PDC
\citep{Stumpe:2012, Smith:2012} \kepler\ data.
Unlike the previous examples, the posterior distribution is sharply
multimodal and na\"ively maximizing the posterior using \project{L-BFGS-B} is
not practical.
Instead, we start by estimating the initial values for $\nu_\mathrm{max}$ and
$\Delta\nu$ using only the month-long subset of data and following the
standard procedure described by \citet{Huber:2009}.
We then run set of \project{L-BFGS-B} optimizations with values of
$\nu_\mathrm{max}$ selected in logarithmic grid centered on our initial
estimate of $\nu_\mathrm{max}$ and initial values of $\Delta\nu$ computed
using the empirical relationship between these two quantities
\citep{Stello:2009}.
This initialization procedure is sufficient for this example, but the general
application of this method will require a more sophisticated prescription.

\Figure{astero} shows a 10~day subset of the dataset used for this example.
The MAP model is overplotted on these data and the residuals away from the
mean prediction of this model are shown in the bottom panel of
\Figure{astero}.
There is no obvious structure in these residuals, lending some credibility to
the model specification.

We initialize \exampleivnwalkers~walkers by sampling from an isotropic
Gaussian centered on the MAP parameters (the full set of parameters and their
priors are listed in Table~\ref{tab:example-4-params}), run
\exampleivnburn~steps of burn-in, and run \exampleivnsteps~steps of MCMC using
\emcee.
We estimate the mean autocorrelation time for the chains of
$\ln\nu_\mathrm{max}$ and $\ln\Delta\nu$ and find \exampleivneff\ effective
samples from the marginalized posterior density.
\Figure{astero-corner} shows the marginalized density for $\nu_\mathrm{max}$
and $\Delta\nu$ compared to the results from the literature.
This result is consistent within the published error bars and the posterior
constraints are tighter than the published results.
All asteroseismic analyses are known to have substantial systematic and
method-dependent uncertainties \citep{Verner:2011} so further experiments
would be needed to fully assess the reliability of this method.

This model requires about 10~CPU minutes to run the MCMC to convergence.
This is more computationally intensive than traditional methods of measuring
asteroseismic oscillations, but it is much cheaper than the same analysis
using a general direct GP solver.
For comparison, we estimate that repeating this analysis using a general
Cholesky factorization implemented as part of a tuned linear algebra
library\footnote{We use the Intel Math Kernel Library
\url{https://software.intel.com/en-us/intel-mkl}} would require about 15~CPU
hours.
An in-depth discussion of the benefits of rigorous probabilistic inference of
asteroseismic parameters in the time domain is beyond the scope of this paper,
but we hope to revisit this opportunity in the future.
}

@@

\subsection{Example 5: Exoplanet transit fitting}

@@

This paper includes data collected by the \kepler\ Mission. Funding for the
\kepler\ Mission is provided by the NASA Science Mission directorate.
We are grateful to the entire \kepler\ team, past and present.
These data were obtained from the Mikulski Archive for Space Telescopes
(MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST is provided by the NASA Office of Space Science via grant
NNX13AC07G and by other grants and contracts.
